# Estimating assumption-free: the world of supervised learning techniques

**Caution: in a highly developmental stage! See Section  \@ref(caution).**

## What machine learning is

What is Machine Learning (ML) (or Statistical Learning)? As the [ISLR book](http://www-bcf.usc.edu/~gareth/ISL/) puts it, it's a "vast set of tools for understanding data". Before we explain more, we need to consider the two main types of ML:

- __Supervised learning__. (_This is the focus of BAIT 509_). Consider a "black box" that accepts some input(s), and returns some type of output. Feed it a variety of input, and write down the output each time (to obtain a _data set_). _Supervised learning_ attempts to learn from these data to re-construct this black box. That is, it's a way of building a forecaster/prediction tool. 

You've already seen examples throughout MBAN. For example, consider trying to predict someone's wage (output) based on their age (input). Using the `Wage` data set from the `ISLR` R package, here are examples of inputs and outputs:

```{r, echo=FALSE}
Wage %>%
    select(age, wage) %>%
    head %>%
    magrittr::set_rownames(1:6)
```

We try to model the relationship between age and wage so that we can predict the salary of a new individual, given their age. 

An example supervised learning technique is _linear regression_, which you've seen before in BABS 507/508. For an age `x`, let's use linear regression to make a prediction that's quadratic in `x`. Here's the fit:

```{r, echo=FALSE}
ggplot(Wage, aes(age, wage)) +
    geom_point(alpha=0.2) +
    geom_smooth(method="lm", se=FALSE,
                formula=y~poly(x,2)) +
    ylab("Wage (1000s of USD)") +
    xlab("Age") +
    theme_bw()
```

The blue curve represents our attempt to "re-construct" the black box by learning from the existing data. So, for a new individual aged 70, we would predict a salary of about \$100,000. A 50-year-old, about \$125,000.


- __Unsupervised learning__. (_BAIT 509 will not focus on this_). Sometimes we can't see the output of the black box. _Unsupervised learning_ attempts to find structure in the data without any output. 

For example, consider the following two gene expression measurements (actually two principal components). Are there groups that we can identify here?

```{r, echo=FALSE}
NCI60[[1]] %>% 
    prcomp() %>% 
    `[[`("x") %>% 
    as_tibble() %>% 
    select(PC1, PC2) %>% 
    ggplot(aes(PC1, PC2)) +
    geom_point(alpha=0.5) +
    ggtitle("Gene expression data") +
    theme_bw()
```

You've seen methods for doing this in BABS 507/508, such as k-means. 



## Types of Supervised Learning

There are two main types of supervised learning methods -- determined entirely by the type of response variable.

- __Regression__ is supervised learning when the response is numeric.
- __Classification__ is supervised learning when the response is categorical. 

We'll examine both equally in this course. 

Note: Don't confuse classification with _clustering_! The latter is an unsupervised learning method.

