---
title: "DSCI 562 Lab 4 Tutorial: GAM"
output: html_document
---

```{r, warning=FALSE, echo=FALSE}
library(knitr)
library(ggplot2)
opts_chunk$set(warning=FALSE, fig.width=5, fig.height=3, fig.align="center")
```

## Generalized Additive Models

To fit a GAM in R, we could use:

1. the function `gam` in the `mgcv` package, or
2. the function `gam` in the `gam` package. 

Differences between the two functions are discussed in the "Details" section of the `gam` documentation in the `mgcv` package. Choose one, but don't load both! `mgcv` tends to be updated more frequently, and is generally more flexible (compare the Index pages), so that's what is used in this tutorial. But the `gam` package has similar workings. 

The `gam` function works similarly to other regression functions, but the formula specification is different. Let's go through different formula specifications, doing regression on the `mtcars` dataset in R.

The formula `mpg ~ disp + wt` gives you a _linear model_. It indicates that `disp` and `wt` both enter the model in a linear fashion.

```{r}
library(mgcv)
fit1 <- gam(mpg ~ disp + wt, data=mtcars)
fit2 <- lm(mpg ~ disp + wt, data=mtcars)
summary(fit1)
summary(fit2)
```

Notice that the coefficient estimates are the same.

To make a term non-parametric, wrap the `s` function around the term (for _splines_; comes with the `mgcv` package). The `gam` package also has a `lo` function, for _loess_ smoothing.

```{r}
fit3 <- gam(mpg ~ s(disp) + s(wt), data=mtcars)
summary(fit3)
```

Now, each predictor enters the model in a non-parametric, additive form. The nonparametric functions can be accessed by calling `plot`. For documentation, see `?plot.gam`. Let's plot the "bivariate" scatterplots behind these curves too (these bivariate data actually use partial residuals).

```{r}
plot(fit3, residuals=TRUE)
```

Looks like the "weight" variable (`wt`) is quite linear. We can let it be linear, while the `disp` variable remains nonparametric. "Wiggliness" of the smoothed fit can be controlled through the `k` argument of the `s` function, but this is chosen in a "smart" way by default.

```{r}
fit4 <- gam(mpg ~ s(disp, k=3) + wt, data=mtcars)
summary(fit4)
plot(fit4, residuals=TRUE)
```

You can even combine predictors into a common smooth function:

```{r}
fit5 <- gam(mpg ~ s(disp, qsec) + wt, data=mtcars)
summary(fit5)
```

For each, the `predict` and `residuals` functions work in the same old way. Let's use them to make a residual plot:

```{r}
qplot(predict(fit5), residuals(fit5)) +
    geom_abline(intercept=0, slope=0, linetype="dashed") +
    xlab("Prediction (mean)") + ylab("Residuals")
```

For their documentation, see `?predict.gam` and `?residuals.gam`.