# GLM's in R

This document introduces the `glm()` function in R for fitting a Generlized Linear Model (GLM). We'll work with the `titanic_train` dataset in the `titanic` package.

```{r}
library(tidyverse)
library(broom)
library(titanic)
dat <- na.omit(titanic_train)
str(dat)
```


Consider the regression of `Survived` on `Age`. Let's take a look at the data with jitter:

```{r, fig.height=2, warning=FALSE}
ggplot(dat, aes(Age, Survived)) +
    geom_jitter(height=0.1, alpha=0.25) +
    scale_y_continuous(breaks=0:1, labels=c("Perished", "Survived")) +
    theme_bw()
```

Recall that the linear regression can be done with the `lm` function:

```{r}
res_lm <- lm(Survived ~ Age, data=dat)
summary(res_lm)
```

In this case, the regression line is ```r res_lm$coefficients[1]``` + ```r res_lm$coefficients[2]``` `Age`.

A GLM can be fit in a similar way, using the `glm` function -- we just need to indicate what type of regression we're doing (binomial? poission?) and the link function. We are doing bernoulli (binomial) regression, since the response is binary (0 or 1); lets choose a `probit` link function.

```{r}
res_glm <- glm(factor(Survived) ~ Age, data=dat, family=binomial(link="probit"))
```

The `family` argument takes a __function__, indicating the type of regression. See `?family` for the various types of regression allowed by `glm()`. 

Let's see a summary of the GLM regression:

```{r}
summary(res_glm)
```

We can make predictions too, but this is not as straight-forward as in `lm()` -- here are the "predictions" using the `predict()` generic function:

```{r}
pred <- predict(res_glm)
qplot(dat$Age, pred) + labs(x="Age", y="Default Predictions")
```

Why the negative predictions? It turns out this is just the linear predictor, ```r res_glm$coefficients[1]``` + ```r res_glm$coefficients[2]``` `Age`.

The documentation for the `predict()` generic function on `glm` objects can be found by typing `?predict.glm`. Notice that the `predict()` generic function allows you to specify the *type* of predictions to be made. To make predictions on the mean (probability of `Survived=1`), indicate `type="response"`, which is the equivalent of applying the inverse link function to the linear predictor.

Here are those predictions again, this time indicating `type="response"`:

```{r}
pred <- predict(res_glm, type="response")
qplot(dat$Age, pred) + labs(x="Age", y="Mean Estimates")
```

Look closely -- these predictions don't actually fall on a straight line. They follow an inverse probit function (i.e., a Gaussian cdf):

```{r}
mu <- function(x) pnorm(res_glm$coefficients[1] + res_glm$coefficients[2] * x)
qplot(dat$Age, pred) + 
    labs(x="Age", y="Mean Estimates") +
    stat_function(fun=mu, colour="blue") +
    scale_x_continuous(limits=c(-200, 200))
```

### `broom::augment()`

We can use the `broom` package on `glm` objects, too. But, just like we had to specify `type="response"` when using the `predict()` function in order to evaluate the model function, so to do we have to specify something in the `broom::augment()` function. Here, the `type.predict` argument gets passed to the `predict()` generic function (actually, the `predict.glm()` method). This means that indicating `type.predict="response"` will evaluate the model function:

```{r}
res_glm %>% 
  augment(type.predict = "response") %>% 
  head()
```

